sdmTMB: Fast, flexible, and user-friendly spatial and spatiotemporal generalized linear mixed-effects models

### 2022-01-07 doc/appendix-model.Rmd
Minor edits

  "$\\omega_{\\s}$", "\\texttt{omega\\_s}", "Spatial random field at point $\\s$ (knot)",
  "$\\omega_{\\s}^*$", "\\texttt{omega\\_s\\_A}", "Spatial random field at point $\\s$ (interpolated)",
  "$\\zeta_{\\s, t}$", "\\texttt{zeta\\_t}", "Spatially varying coefficient random field at point $\\s$ and time $t$ (knot)",
  "$\\zeta_{\\s, t}^*$", "\\texttt{zeta\\_t\\_A}", "Spatially varying coefficient random field at point $\\s$ and time $t$ (interpolated)",
  "$\\epsilon_{\\s, t}$", "\\texttt{epsilon\\_st}", "Spatiotemporal random field at point $\\s$ and time $t$ (knot)",
  "$\\epsilon_{\\s, t}^*$", "\\texttt{epsilon\\_st\\_A}", "Spatiotemporal random field at point $\\s$ and time $t$ (interpolated)",
  "$\\delta_{\\s,t}$", "\\texttt{b\\_t}", "AR1 or random walk spatiotemporal deviations (knot)",
  "$\\alpha_g$", "\\texttt{RE}", "IID random intercept deviation for group $g$",

@@

The complete sdmTMB model can be written as:

@@

Internally, the random fields follow a Gaussian Markov random field (GMRF):

$$
\boldsymbol{\omega} \sim \mathrm{MVNormal}\left(\boldsymbol{0}, \sigma_\omega^2 \boldsymbol{Q}^{-1}_\omega\right)
$$
where $\boldsymbol{Q}_\omega$ is a sparse precision matrix and $\sigma_\omega^2$ is the marginal variance.

@@

Spatiotemporal random fields are included by default if there are multiple time elements (`time` argument is not `NULL`) and can be set to IID (`'iid'`; default), AR1 (`'ar1'`), random walk (`'rw'`), or off (`'off'`) via the `spatiotemporal` argument.
These text values are case insensitive.

@@

Similarly to the spatial random fields, these spatiotemporal random fields (including all versions described below) are parameterized internally with a sparse precision matrix ($\boldsymbol{Q}_\epsilon$):

@@

The random fields are parameterized internally with a sparse precision matrix ($\boldsymbol{Q}_\zeta$):

@@

where $\mu$ is the mean, $p$ is the power parameter constrained between 1 and 2, and $\phi$ is the dispersion parameter.
The Tweedie distribution can be helpful for modelling data that are positive and continuous but also contains zeros.

@@

The values of the spatial variables at the knots are multiplied by a projection matrix $\boldsymbol{A}$ that bilinearly interpolates from the knot locations to the values at the locations of the observed or predicted data [@lindgren2015]:

$$
\boldsymbol{\omega}^* = \boldsymbol{A} \boldsymbol{\omega},
$$
where $\boldsymbol{\omega}^*$ represents the values of the spatial random fields at the observed locations or predicted data locations.
The matrix $\boldsymbol{A}$ has a row for each data point or prediction point and a column for each knot.
Three non-zero elements on each row define the weight of the neighbouring 3 knot locations for location $s$.
The same bilinear interpolation happens for any spatiotemporal random fields and $\boldsymbol{\epsilon}_t$:

>>>>

  "$\\omega_{\\s}$", "\\texttt{omega\\_s}", "Spatial random field at point $\\s$ (knot)",
  "$\\omega_{\\s}^*$", "\\texttt{omega\\_s\\_A}", "Spatial random field at point $\\s$ (interpolated)",
  "$\\zeta_{\\s, t}$", "\\texttt{zeta\\_t}", "Spatially varying coefficient random field at point $\\s$ and time $t$ (knot)",
  "$\\zeta_{\\s, t}^*$", "\\texttt{zeta\\_t\\_A}", "Spatially varying coefficient random field at point $\\s$ and time $t$ (interpolated)",
  "$\\epsilon_{\\s, t}$", "\\texttt{epsilon\\_st}", "Spatiotemporal random field at point $\\s$ and time $t$ (knot)",
  "$\\epsilon_{\\s, t}^*$", "\\texttt{epsilon\\_st\\_A}", "Spatiotemporal random field at point $\\s$ and time $t$ (interpolated)",
  "$\\delta_{\\s,t}$", "\\texttt{b\\_t}", "AR(1) or random walk spatiotemporal deviations (knot)",
  "$\\alpha_g$", "\\texttt{RE}", "IID random intercept deviation for group $g$",

@@

The complete sdmTMB model can be written as

@@

Internally, the random fields follow a Gaussian Markov random field (GMRF)

$$
\boldsymbol{\omega} \sim \mathrm{MVNormal}\left(\boldsymbol{0}, \sigma_\omega^2 \boldsymbol{Q}^{-1}_\omega\right),
$$
where $\boldsymbol{Q}_\omega$ is a sparse precision matrix and $\sigma_\omega^2$ is the marginal variance.

@@

Spatiotemporal random fields are included by default if there are multiple time elements (`time` argument is not `NULL`) and can be set to IID (independent and identically distributed, `'iid'`; default), AR(1) (`'ar1'`), random walk (`'rw'`), or off (`'off'`) via the `spatiotemporal` argument.
These text values are case insensitive.

@@

Similarly to the spatial random fields, these spatiotemporal random fields (including all versions described below) are parameterized internally with a sparse precision matrix ($\boldsymbol{Q}_\epsilon$)

@@

The random fields are parameterized internally with a sparse precision matrix ($\boldsymbol{Q}_\zeta$)

@@

where $\mu$ is the mean, $p$ is the power parameter constrained between 1 and 2, and $\phi$ is the dispersion parameter.
The Tweedie distribution can be helpful for modelling data that are positive and continuous but also contain zeros.

@@

The values of the spatial variables at the knots are multiplied by a projection matrix $\boldsymbol{A}$ that bilinearly interpolates from the knot locations to the values at the locations of the observed or predicted data [@lindgren2015]

$$
\boldsymbol{\omega}^* = \boldsymbol{A} \boldsymbol{\omega},
$$
where $\boldsymbol{\omega}^*$ represents the values of the spatial random fields at the observed locations or predicted data locations.
The matrix $\boldsymbol{A}$ has a row for each data point or prediction point and a column for each knot.
Three non-zero elements on each row define the weight of the neighbouring 3 knot locations for location $\s$.
The same bilinear interpolation happens for any spatiotemporal random fields

### 2022-01-07 doc/paper.Rmd
Minor edits

---
title: "sdmTMB: Fast, flexible, and user-friendly spatial and spatiotemporal generalized linear mixed-effects models"
author: "|
  Sean C. Anderson$^1$^[Corresponding author: sean.anderson@dfo-mpo.gc.ca],
  Eric J. Ward$^2$,
  Philina A. English$^1$,
  Lewis A. K. Barnett$^3$\n|
  $^1$Pacific Biological Station, Fisheries and Oceans Canada, Nanaimo, BC, Canada\n|
  $^2$Northwest Fisheries Science Center, National Marine Fisheries Service,\n|
      National Oceanographic and Atmospheric Administration, Seattle, WA, USA\n|
  $^3$Alaska Fisheries Science Center, National Marine Fisheries Service,\n|
      National Oceanographic and Atmospheric Administration, Seattle, WA, USA\n"
output:
    # bookdown::word_document2:
    bookdown::pdf_document2:
      toc: false
      number_sections: false
      fig_caption: true
      highlight: "monochrome"
csl: mee.csl
bibliography: refs.bib
link-citations: yes
header-includes:
  \usepackage{setspace}\onehalfspacing
  \usepackage[left]{lineno}
  \usepackage{bm}
  \usepackage{amssymb}
  \usepackage{tablefootnote}
  \linenumbers
  \newcommand{\beginsupplement}{
  \setcounter{equation}{0}
  \renewcommand{\theequation}{S.\arabic{equation}}
  \setcounter{table}{0}
  \renewcommand{\thetable}{S\arabic{table}}
  \setcounter{figure}{0}
  \renewcommand{\thefigure}{S\arabic{figure}}}
  \widowpenalty10000
  \clubpenalty10000
editor_options:
  markdown:
    wrap: sentence
abstract: |
    | \noindent 1. Geostatistical data---spatially referenced observations related to some continuous spatial phenomenon--are ubiquitous in ecology and can reveal ecological processes and inform management decisions through applications such as species distribution models (SDMs). However, appropriate models to analyze such data, such as generalized linear mixed effects models (GLMMs) with Gaussian random fields, are often computationally intensive and challenging to implement, interpret, and evaluate.
    | 2. Here, we introduce the R package sdmTMB, which implements predictive-process SPDE- (stochastic partial differential equation) based spatial and spatiotemporal models. Estimation is conducted via maximum marginal likelihood with Template Model Builder (TMB), but can be extended to penalized likelihood or Bayesian inference. We describe the statistical model, illustrate the package's use through two case studies, and compare the functionality, speed, and interface to related software.
    | 3. We highlight advantages of using sdmTMB for this class of models: (1) sdmTMB provides a flexible interface familiar to users of `glm()`, lme4, glmmTMB, or mgcv; (2) estimation is often faster than alternatives; (3) sdmTMB provides simple out-of-sample cross validation; (4) non-stationary processes (time-varying and  spatially varying coefficients) are easily constructed with a formula interface; and (5) sdmTMB includes features not available as a combination in related packages (e.g., penalized smoothers and break-point effects, anisotropy, abundance index standardization).
    | 4. We hope that sdmTMB's user-friendly interface will open this useful class of models to a wider audience within species distribution modelling and beyond.
    | Keywords: Gaussian Markov random fields, generalized linear mixed effects models, INLA, species distribution modelling, R package, spatio-temporal, spatial-temporal, Template Model Builder
---

>>>>

---
title: "sdmTMB: Fast, flexible, and user-friendly spatial and spatiotemporal generalized linear mixed-effects models"
author: "|
  Sean C. Anderson$^1$^[Corresponding author: sean.anderson@dfo-mpo.gc.ca],
  Eric J. Ward$^2$,
  Philina A. English$^1$,
  Lewis A. K. Barnett$^3$\n|
  $^1$Pacific Biological Station, Fisheries and Oceans Canada, Nanaimo, BC, Canada\n|
  $^2$Northwest Fisheries Science Center, National Marine Fisheries Service,\n|
      National Oceanographic and Atmospheric Administration, Seattle, WA, USA\n|
  $^3$Alaska Fisheries Science Center, National Marine Fisheries Service,\n|
      National Oceanographic and Atmospheric Administration, Seattle, WA, USA\n"
output:
    # bookdown::word_document2:
    bookdown::pdf_document2:
      toc: false
      number_sections: false
      fig_caption: true
      highlight: "monochrome"
csl: mee.csl
bibliography: refs.bib
link-citations: yes
header-includes:
  \usepackage{setspace}\onehalfspacing
  \usepackage[left]{lineno}
  \usepackage{bm}
  \usepackage{amssymb}
  \usepackage{tablefootnote}
  \linenumbers
  \newcommand{\beginsupplement}{
  \setcounter{equation}{0}
  \renewcommand{\theequation}{S.\arabic{equation}}
  \setcounter{table}{0}
  \renewcommand{\thetable}{S\arabic{table}}
  \setcounter{figure}{0}
  \renewcommand{\thefigure}{S\arabic{figure}}}
  \widowpenalty10000
  \clubpenalty10000
editor_options:
  markdown:
    wrap: sentence
abstract: |
    | \noindent 1. Geostatistical data---spatially referenced observations related to some continuous spatial phenomenon---are ubiquitous in ecology and can reveal ecological processes and inform management decisions through applications such as species distribution models (SDMs). However, appropriate models to analyze such data, such as generalized linear mixed effects models (GLMMs) with Gaussian random fields, are often computationally intensive and challenging to implement, interpret, and evaluate.
    | 2. Here, we introduce the R package sdmTMB, which implements predictive-process SPDE- (stochastic partial differential equation) based spatial and spatiotemporal models. Estimation is conducted via maximum marginal likelihood with Template Model Builder (TMB), but can be extended to penalized likelihood or Bayesian inference. We describe the statistical model, illustrate the package's use through two case studies, and compare the functionality, speed, and interface to related software.
    | 3. We highlight advantages of using sdmTMB for this class of models: (1) sdmTMB provides a flexible interface familiar to users of `glm()`, lme4, glmmTMB, or mgcv; (2) estimation is often faster than alternatives; (3) sdmTMB provides simple out-of-sample cross validation; (4) non-stationary processes (time-varying and  spatially varying coefficients) are easily constructed with a formula interface; and (5) sdmTMB includes features not available as a combination in related packages (e.g., penalized smoothers and break-point effects, anisotropy, abundance index standardization).
    | 4. We hope that sdmTMB's user-friendly interface will open this useful class of models to a wider audience within species distribution modelling and beyond.
    | Keywords: Gaussian Markov random fields, generalized linear mixed effects models, INLA, species distribution modelling, R package, spatio-temporal, spatial-temporal, Template Model Builder
---
